```{r setup, include=FALSE}
library(knitr)
library(lubridate)
library(dplyr)
library(ggplot2)
library(tidyr)
opts_chunk$set(cache=TRUE)
```
Downloading and reading the storm data
```{r download, cache=TRUE}
url = "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(url, "stormData.csv.bz2")
stormData <- read.csv("stormData.csv.bz2", na.strings = "")
```
Changing date to a poxit variable (not neccessary for non-exploratory analysis).
Tidy variable names (lower case, no dots whitespace or underscores. 
```{r processing data}
data <- stormData
data$BGN_DATE <- as.Date(data$BGN_DATE, format="%m/%d/%Y %H:%M:%S")
data$END_DATE <- as.Date(data$END_DATE, format = "%m/%d/%Y %H:%M:%S")

data <- mutate(data, year=year(BGN_DATE))
# Tidying the names, so lower case and no "_/ /."
names(data) <- tolower(names(data))
names(data) <- gsub("_", "x", names(data))
```

Events span 50 years. Over this period collection method has changed substantially 
  - *1950:1954:* Only Tornados were recorded. 
  - *1955:1992:* Only tornado, thunderstorm, wind and hail events were keyed from paper publications into digital. 
  - *1993:1995:* Only tornado, thuderstorm, wind and hail events have been extracted from the unformated text files. 
  - *1996:2011:* All 48 event types recorded
  
Over time the number and type of recorded events have incresed. 
```{r recorded events with time}
years <- mutate(data, year= year(bgnxdate)) %>%
  group_by(year) %>%
  summarise(number=n())
       
filter(years, year %in% c(1950,1960, 1970, 1980, 1990, 2000, 2010))
```

event types
====================

Although the documenation says there should only be 48 levels to EVTYPE there are in fact 985
```{r evtype levels}
length(levels(as.factor(data$evtype)))
```

Trying to match the evtypes: 
first look at distribution of evtypes. 
```{r grouping levels}
evtype.levels <- group_by(data, evtype) %>% 
  summarise(number = n()) %>%
  arrange(desc(number))
head(evtype.levels)
```

Defining a list of allowed evtypes from the documentation, we want to match as many of the observations to these 48 data types as possible. 
```{r listing evtype}
eventNames <- c("Astronomical Low Tide", "Avalanche", "Blizzard", "Coastal Flood", "Cold/Wind Chill", "Debris Flow", "Dense Fog", "Dense Smoke", "Drought", "Dust Devil", "Dust Storm", "Excessive Heat", "Extreme Cold/Wind Chill", "Flash Flood", "Flood", "Frost/Freeze", "Funnel Cloud", "Freezing Fog", "Hail", "Heat", "Heavy Rain", "Heavy Snow", "High Surf", "High Wind", "Hurricane (Typhoon)", "Ice Storm", "Lake-Effect Snow", "Lakeshore Flood", "Lightning", "Marine Hail", "Marine High Wind", "Marine Strong Wind", "Marine Thunderstorm Wind", "Rip Current", "Seiche", "Sleet", "Storm Surge/Tide", "Strong Wind", "Thunderstorm Wind", "Tornado", "Tropical Depression", "Tropical Storm", "Tsunami", "Volcanic Ash", "Waterspout", "Wildfire", "Winter Storm", "Winter Weather") 
```

Normalising all entries to lower case, two thirds of the data matches these event names. 
```{r matching event names}
eventNames <- tolower(eventNames)
data$evtype <- tolower(data$evtype)
matches <- sapply(data$evtype, function(x) x %in% eventNames)
sum(matches) / nrow(data)
matched <- data[matches,]
```

Looking at the subset of evtype levels which are not matched to the 48 allowed classifications:  
96% of them are contained in the 10 most common entries. 
```{r looking at not matched data}
not.matched <- group_by(data[!matches,], evtype) %>%
                       summarise(number = n()) %>%
                       arrange(desc(number))

sum(not.matched[1:10,2]) / sum(not.matched[,2])
```

None of the correct classifications end in an "s" but many of the incorrect classifications are just plural (...s) versions of correct classifications e.g. winds, storms, currents, floods, squalls (even windss) so remove trailing "s" characters evtypes and then recalculate matched statistics. Improved from 635351 to 658483 (~2.5% more of the entries now match)
N.B. this can cause errors if a new correct classification is added ending with an "s", to fix this you could apply this procedure only to non-matching evtypes, and restrict to only removine one "s". 
```{r removing plural evtypes}
data$evtype <- sapply(data$evtype, function(x) gsub("s+$", "", x))

matches <- sapply(data$evtype, function(x) x %in% eventNames)
sum(matches) / nrow(data)
```

Manually reallocating the most popular non-matched evtypes. 
```{r reallocating evtypes}
# tstm wind -(stands for too small to measure) this puts it in the strong wind category (as long as some injury or damage or fatality was suffered. (p.s. having the requirement of damage for definition will skew the analysis to say low speed wind is dangerous!)
# Note the majority of the tstm winds did not have injuries/fatalities/damage so were left alone
data$evtype <-
  ifelse(grepl("tstm wind", data$evtype) & ((data["fatalities"] > 0) | (data["injuries"] > 0) | (data["propdmg"] > 0) | (data["cropdmg"] > 0)),
         sub("tstm wind", "strong wind", data$evtype), 
         as.character(data$evtype))

# marine tstm wind == marine strong wind (as long as injury / damage / fatality > 0)
data$evtype <-
  ifelse(grepl("marine tstm wind", data$evtype) & ((data["fatalities"] > 0) | (data["injuries"] > 0) | (data["propdmg"] > 0) | (data["cropdmg"] > 0)),
       sub("marine tstm wind", "strong wind", data$evtype), 
       as.character(data$evtype))

# urban/sml stream fld  (hard to tell without additional information weather this category refers to flooding or flash flooding)
  # - flash floods do not exist for more than 2 to 3 days
  # - table floods / flash floods and default to most common. 
data$evtype <-
  ifelse((data$evtype == "urban/sml stream fld") & ((data$endxdate - data$bgnxdate) <= 3),
         sub("urban/sml stream fld", "flash flood", data$evtype),
         as.character(data$evtype))
data$evtype <-
  ifelse((data$evtype == "urban/sml stream fld") & ((data$endxdate - data$bgnxdate) <= 3),
         as.character(data$evtype), 
         sub("urban/sml stream fld", "flood", data$evtype))

# wild/forest fire  == wildfire
data$evtype <- gsub("wild/forest fire", "wildfire", data$evtype)
 
# winter weather/mix  = winter weather
data$evtype <- gsub("winter weather/mix", "winter weather", data$evtype)

# tstm wind/hail  no real way to categorise this into one category so left alone

# flash flooding == flash flood
data$evtype <- gsub("flash flooding", "flash flood", data$evtype)

# extreme cold  = extreme cold/wind chill
data$evtype <- gsub("extreme cold.*", "extreme cold/wind chill", data$evtype)

# flood/flash flood  (as long as flood is less than or equal to 3 days, assume flash flood, this is because flash floods are far more common ~2x as likely and is otherwise up to recorders discression (no finite boundaries between the two)
data$evtype <-
  ifelse((data$evtype == "flood/flash flood") & ((data$endxdate - data$bgnxdate) <= 3),
         as.character(data$evtype), 
         sub("flood/flash flood", "flood", data$evtype))
data$evtype <-
  ifelse((data$evtype == "flood/flash flood") & ((data$endxdate - data$bgnxdate) <= 3),
         sub("flood/flash flood", "flood", data$evtype),
         as.character(data$evtype))

data$evtype <- 
  ifelse( grepl("flood", data$evtype) & !grepl("flash", data$evtype), 
          "flood",
          as.character(data$evtype))
# snow  (heavy snow is only snow related category, definition of heavy snow is regionally defined so difficult to categorise with data availible) so leave this variable alone

# hail is labelled by its size, remove size labelling from the end of hail
data$evtype <- sub("^hail [0-9]*", "hail", data$evtype)

# landslide / mudslide are categorised as debris flow
data$evtype <- sub("landslide", "debris flow", data$evtype)
data$evtype <- sub("mudslide", "debris flow", data$evtype)

# hurricane (typhoon) often labelled as "hurricane" or "hurricane/typhoon"
data$evtype <- sub("hurricane", "hurricane (typhoon)", data$evtype)
data$evtype <- sub("typhoon", "hurricane (typhoon)", data$evtype)

```

Approximately 80% (80.7%) of the data is now categorised within the 48 allowed factors. However of those that are left 90% of it is due to too small the measure wind which caused no injury or damage. These don't fall in any category and have been ignored on purpose. 


analysis
==========

### Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health

Harm can be categorised both as injury and fatality. It is impossible to definitively compare these two so we shall look at them seperately. 

We will only look at the observations with allowed evtypes after trying to match these as effectively as possible. 

As 1996 was the [first year all 48 event types were recorded](https://www.ncdc.noaa.gov/stormevents/details.jsp) we will analyse from only this time 
```{r selecting relavent subset}
subset <- select(data, bgnxdate, endxdate, year, evtype, fatalities, injuries, propdmg, propdmgexp, cropdmg, cropdmgexp, remarks, wfo, refnum) %>%
  filter(year > 1995)
subset <- subset[subset$evtype %in% eventNames,]
subset$evtype <- factor(subset$evtype) # resets the levels
```

While we have accounted for the increased types of events recorded there is still a correlation with number of observations and time. As this histogram binning each year shows. 

```{r another observation time histogram}
qplot(bgnxdate, data = subset, binwidth = 365, ylim = c(0,80000))
```
This is likeley due to improved recording over time. One way to factor for this is to produce a weighting based ona  linear regression model over time. The difficulty here is that certain event types will be affected more/less by this record keeping, for example tornadoes which have been recorded since 1950 are probably quite acccurately recorded or tsunamis which are very devistating (and hard to miss) are also probably accurately recorded. However if you complete seperate linear regression models for each event type you would hide any time related phenomenon which could be occuring. Also some event types are rare and there would be too little data to fit a linear model. Because of these we  will take a single linear regression model for all event types but be aware that events such as tornadoes or tsunamis are likely over estimates.  

```{r linear regression with time}
yearly.observations <- group_by(subset, year) %>%
  summarise(events = n())

# We use a linear model instead of just relating observations as a proportion of acutal 2011 readings so that the system is "smoothed", if 2011 is slightly above/below average it doesn't have a big effect and if the year you are weighting had a lot of tornadoes that doesn't have a large effect on how heat waves are weighted that year. 
year.model <- lm(events ~ year, yearly.observations)

years <- yearly.observations$year

year.predict <- predict(year.model, data.frame(year = seq(1996,2011,1)))
yearly.observations$lmpredict <- year.predict
yearly.observations <- mutate(yearly.observations, weighting = lmpredict[length(lmpredict)] / lmpredict, weighted.events = events*weighting)
qplot(year, weighted.events, data=yearly.observations, ylim=c(0,80000)) + geom_smooth(method="lm")
```

As you can see from the plot, there are some outlier years, 1996,1998, 2008,2011 where there were a lot of observed weather events but the majority of years are now flatly spread. 

```{r weighting injures/fatalities}
evtype.health <- group_by(subset,year, evtype) %>%
  summarise(injuries = sum(injuries), fatalities = sum(fatalities))

weighted.injuries <- select(evtype.health, -fatalities) %>%
  spread(evtype, injuries)
weighted.injuries[,2:length(weighted.injuries)] <- apply(weighted.injuries[,2:length(weighted.injuries)], 2, function(x) x*yearly.observations$weighting)
weighted.injuries <- gather(weighted.injuries, evtype, injuries, -year)

weighted.fatalities <- select(evtype.health, -injuries) %>%
  spread(evtype, fatalities)
weighted.fatalities[,2:length(weighted.fatalities)] <- apply(weighted.fatalities[,2:length(weighted.fatalities)], 2, function(x) x*yearly.observations$weighting)
weighted.fatalities <- gather(weighted.fatalities, evtype, fatalities, -year)

weighted.evtype.health <- weighted.injuries
weighted.evtype.health$fatalities <- weighted.fatalities$fatalities
```
As seen in the plot, tornadoes are the biggest cause of documented injuries and fatalities. I am still potentially skeptical of this due tornadoes likely being well documented in contrast to the other evtypes, and consistently documented (leading to the linear model inflating it disproportionately. That said even as an overestimate tornadoes are a large concern. With flooding and excessive heat as the two other main causes of injury.  
Excessive heat is a much bigger cause of fatality, and flash floods are a much bigger concern than floods in this regard. While tornadoes are still a major concern. 
```{r plotting high fatality or injury event types}
total.injuries <- group_by(weighted.evtype.health, evtype) %>%
  summarise(injuries = sum(injuries, na.rm=T), fatalities = sum(fatalities, na.rm=T)) %>%
  arrange(desc(injuries))

temp <- head(arrange(total.injuries, desc(injuries)),10)
qplot(evtype, injuries, data=temp)

temp <- head(arrange(total.injuries, desc(fatalities)),10)
qplot(evtype, fatalities, data=temp)
```

## Across the United States, which types of events have the greatest economic consequences?

Property damage and crop damage are combinations of two numbers, a number and a multiplier. The allowed values for the multiplier are "B" "M" and "K" for billion, million, kilo(thousand). In the total dataset, there are a number of observations outside of these values. However in the subset of 1996+ and the converted event type factors there is only one observation outside these factors. This observation has a prop multiplier of 0 but also a propdmg value of 0, so the 0 can be changed to an NA. 
```{r replacing unallowed values}
filter(subset, propdmgexp == "0")
subset$cropdmgexp <- factor(subset$cropdmgexp)
subset$propdmgexp <- factor(subset$propdmgexp)
subset$propdmgexp[which(subset$propdmgexp == "0")] <- NA
```

We now multiply the values of cropdmg and propdmg by the multipliers 1000 for K etc. 
```{r calculating prop and crop damage}
if (subset$cropdmgexp == "K") { subset$cropdamage <- subset$cropdmg * 1000
} else if (subset$cropdmgexp == "M") { subset$cropdamage <- subset$cropdmg * 1000000
} else if (subset$cropdmgexp == "B") { subset$cropdamage <- subset$cropdmg * 1000000000
} else { subset$cropdamage <- NA
}

if (subset$propdmgexp == "K") {subset$propdamage <- subset$propdmg * 1000
}else if (subset$propdmgexp == "M") {subset$propdamage <- subset$propdmg * 1000000
}else if (subset$propdmgexp == "B") {subset$propdamage <- subset$propdmg * 1000000000
} else {subset$propdamage <- NA
}
```

```{r weighting damage}
damage.analysis <- select(subset, evtype, cropdamage, propdamage, year) %>%
  group_by(evtype, year) %>%
  summarise(cropdamage = sum(cropdamage), propdamage = sum(propdamage))

weighted.cropdmg <- select(damage.analysis, -propdamage) %>%
  spread(evtype, cropdamage)
weighted.cropdmg[,2:length(weighted.cropdmg)] <- apply(weighted.cropdmg[,2:length(weighted.cropdmg)], 2, function(x) x*yearly.observations$weighting)
weighted.cropdmg <- gather(weighted.cropdmg, evtype, cropdamage, -year)

weighted.propdmg <- select(damage.analysis, -cropdamage) %>%
  spread(evtype, propdamage)
weighted.propdmg[,2:length(weighted.propdmg)] <- apply(weighted.propdmg[,2:length(weighted.propdmg)], 2, function(x) x*yearly.observations$weighting)
weighted.propdmg <- gather(weighted.propdmg, evtype, propdamage, -year)

weighted.damage <- weighted.cropdmg
weighted.damage$propdamage <- weighted.propdmg$propdamage
```

```{r damage plots}
test <- group_by(weighted.damage, evtype) %>%
  summarise(cropdamage = sum(cropdamage, na.rm=T), propdamage = sum(propdamage, na.rm=T))

temp <- arrange(test, desc(cropdamage))
qplot(evtype,  cropdamage, data=head(temp,10))

temp <- arrange(test, desc(propdamage))
qplot(evtype, propdamage, data=head(temp,10))
```